---
title: "Predictions on Movie Ratings"
author: "André Jakob"
date: "12/06/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE }

library(tidyverse)
library(knitr)
#First I have loaded the pre-whriten code given by the course. It installs the needed packages and mount the dataset.

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)


ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

dim(ratings)
movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
dim(movies)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
dim(movielens)
ls()
# Validation set will be 10% of MovieLens data

set.seed(1) # if using R 3.6.0: set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)

edx <- movielens[-test_index,]

temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

#Beginning knowing the amount of rows and columns of the dataset.
dim(edx)
#Show the amount each rate.
count(edx,rating)
#Show the amount of differente movies and users.
names(edx)

```
## Introduction

This report is part of the professional certificate program in Data Science offered by Edx and HarvardX.
Since this is evaluated by other students, I would like to ask the evaluators to be kind with my grammatical errors, since english is not my native language. So I thank you in advance, both for the patience and the evaluation itself.
The purpose of the exercise is to write a Mashine Learning algorithim that predicts the potential movie rating from the given ratings dataset.

## Method
In this chapter I describe the step of the work done, so it has basically the same content as the code writen to acchieve the prediction. Obviously you can read both, or just one of those, what you think is better. Generally I tried to write each step done before the respective code, so if you prefer read the steps right among the codes, please go to the next chapter.
We started from the movielens dataset, which has an enormous dimmension of 10,000,054 observations. After some previous analysis on the data to know how it was structured, we followed the lessons of the previous EdX courses, which teached that are some steps to be done to acchieve a good prediction from a great data set as these.
First we defined a sample index with 10% of the size of the original data set. This sample was divided between test and training set. Below we see how many users and movies the training set has.
``` {r summary train set, echo=TRUE }
k1<-edx %>%
  summarize(n_users = n_distinct(userId),
            n_movies = n_distinct(movieId))
kable(k1)
```


By the way, an important background to this exercise is to know that each user can rate the movies from 0.5 to 5, in 0.5 intervals. Even so, we see in the graphic below that a fractional rating is more rare than a rounded one.
```{r, include=FALSE, echo=FALSE }
#See whose are the most rated movies.
tableratings<-edx %>%summarize(n_users = n_distinct(userId),n_movies = n_distinct(movieId))
table(tableratings)
#And whit ratings are more commom.
count(edx,rating)
#In the graphic below we see that the users tend to give rounded ratings more than fractionated ones.

```

```{r graphic ratings frequence, echo=TRUE }
edx %>%
  group_by(rating) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = rating, y = count)) +
  geom_line()
```

The training set was used to train an algorithm and respectively predictions. So those predictions were applied in the test set. 
The general idea was to start with the simplest possible prediction, and then adding complexions. Each improvement went along with the measurement of its Root-mean-square deviation - a measurement analog as the standard deviation. 
The simplest possible prediction was using just the average,  regardless of the users and movies.
As expected,we had a too high RMSE for about to one, which is obviously not enough, since the possible ratings go from 0.5 to 5. Using just the average to predict the ratings, regardless of the criteria, is almost good enough as guessing blindly. 
The next step was then add our first bias: the movies. First we used the least squares method to estimate this bias. Since the whole data-set is enormous, we therefore estimated this value from our train data set. 
With one bias the prediction was a little better, not good enough though. So we added another bias for users. Each user rates the movies according to their own criteria. Some users tend to give better ratings while others are more rigorous. Below I show a histogram for the average of stars each user gives. There is a substantial variability among them.
```{r, include=FALSE, echo=FALSE }
#Defining previously the RMSE (obviously I will perfect this further)
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
#The first and most basical approach is predict a rating regardless of the user and movie.
mu_hat<-mean(edx$rating)
mu_hat
naive_rmse <- RMSE(validation$rating, mu_hat)
naive_rmse
predictions <- rep(3, nrow(validation))
RMSE(validation$rating, predictions)
rmse_results <- data_frame(method = "Just the average", RMSE = naive_rmse)
rmse_results
#as expected,we have a RMSE for about to one, which is obviously not enough, since the possible ratings go from 0.5 to 5.
#Using just the average to predict the ratings, regardless of the criterias is almost good enough as guessing blindly. The first thing we will compute now is the bias for movies. 
#First we use the least squares method to estimate this bias. Since the whole dataset is enourmous, we therefore estimate this value from our train data set. 
mu <- mean(edx$rating) 
movie_avgs <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu))

predicted_ratings <- mu + validation %>% 
  left_join(movie_avgs, by='movieId') %>%
  pull(b_i)

model_1_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie Effect Model",  
                                     RMSE = model_1_rmse))
rmse_results
#This time the prediction is a little better, not good enough though.
#Another bias is the user. Each user rates the movies according to their own criteria. Some users tend to give better ratings while others are more rigorous. Below I show a histogram for the average of stars each user gives. There is a substantial variability among them.

```


``` {r histogram rating average per user, echo=TRUE }
edx %>% 
  group_by(userId) %>% 
  summarize(b_u = mean(rating)) %>% 
  filter(n()>=100) %>%
  ggplot(aes(b_u)) + 
  geom_histogram(bins = 30, color = "black")
```

For the same reason said above, we could not run a prediction for the whole data set, so we estimated the bias from the training set instead and ran the prediction, achieving a prediction up to 0.865.
Even so, we took a look at the greatest mistakes to learn what could be done to make the algorithm even better. We learned that some obscure movies can mess the prediction. So we saw the top and worse 10 movies according to the prediction and how often they were rated. Below both tables.

```{r, include=FALSE, echo=FALSE }
#For the same reason said above, we cannot run a prediction for the whole data set, so I estimate the bias from the training set instead.
user_avgs <- edx %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
#And run the prediction to see if it have improved.
predicted_ratings <- validation %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)


model_2_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie + User Effects Model",  
                                     RMSE = model_2_rmse))
rmse_results
#It has, to 0.865.
#Even so, I will take a look at the greatest mistakes.
validation %>% 
  left_join(movie_avgs, by='movieId') %>%
  mutate(residual = rating - (mu + b_i)) %>%
  arrange(desc(abs(residual))) %>% 
  select(title,  residual) %>% slice(1:10)
#To avoid obscure movies to mess the prediction
#Let's see the top and worse 10 movies according to the prediction.
movie_titles <- movielens %>% 
  select(movieId, title) %>%
  distinct()

```

```{r top 10 movies, echo=TRUE }
k2<-movie_avgs %>% left_join(movie_titles, by="movieId") %>%
  arrange(desc(b_i)) %>% 
  select(title, b_i) %>% 
  slice(1:10)
kable(k2)
```

```{r worse 10 movies, echo=TRUE }
k3<-movie_avgs %>% left_join(movie_titles, by="movieId") %>%
  arrange(b_i) %>% 
  select(title, b_i) %>% 
  slice(1:10)
kable(k3)
```  
 
To avoid obscure movies influence our prediction, we checked how many ratings those movies became.Below both tables.

```{r number of ratings between top 10 movies, echo=TRUE }
#Lets see how often the were rated.
k4<-edx %>% count(movieId) %>% 
  left_join(movie_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(desc(b_i)) %>% 
  select(title, b_i, n) %>% 
  slice(1:10)
kable(k4)
```

```{r number of ratings between worse 10 movies, echo=TRUE }
k5<-edx %>% count(movieId) %>% 
  left_join(movie_avgs) %>%
  left_join(movie_titles, by="movieId") %>%
  arrange(b_i) %>% 
  select(title, b_i, n) %>% 
  slice(1:10)
kable(k5)
```

As expected, most of them have been rated just a few times. To avoid this, it is common to use a method of regularization of the data. In this case a proper way is to penalize some of the ratings. The penalty therm is called lambda and before using it we needed to calculate the better value using cross validation. Below the plotted lambdas tested.

```{r include=FALSE,echo=FALSE}
#As expected, most of them have been rated just a few times. To avoid this, It is common to use a method of regularization of the data. In this case a proper way is to penalize some of the ratings. The penalty therm is called lambda. First, I need to find the better lambda value using cross validation.
lambdas <- seq(0, 10, 0.25)

rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  return(RMSE(predicted_ratings, validation$rating))
})
lambda <- lambdas[which.min(rmses)]
lambda
#Below a plot to see the lambdas and the minimum.
```

```{r plot lambdas, echo=TRUE }
qplot(lambdas, rmses)
```

We founded the 5.25 as the better value and then ran the prediction using this lambda.
We compared the last method with the other calculations and, as the table shows, after predicting the regularized Movie and user effect model we got the minimum RMSE of 0.86. 

## RMSE

The table below gives the RMSE founded using different complexions of calculations. The prediction with regularized movie and user effect model has got the minimum RMSE of 0.8648170.
```{r include=FALSE,echo=FALSE}
#...Which is 5.25. Let's run the prediction using this lambda.
mu <- mean(edx$rating)
movie_reg_avgs <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = sum(rating - mu)/(n()+lambda), n_i = n())

data_frame(original = movie_avgs$b_i, 
           regularlized = movie_reg_avgs$b_i, 
           n = movie_reg_avgs$n_i) %>%
  ggplot(aes(original, regularlized, size=sqrt(n))) + 
  geom_point(shape=1, alpha=0.5)

predicted_ratings <- validation %>% 
  left_join(movie_reg_avgs, by = "movieId") %>%
  mutate(pred = mu + b_i) %>%
  pull(pred)

model_3_rmse <- RMSE(predicted_ratings, validation$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie Effect Model",  
                                     RMSE = model_3_rmse))

#...and compare with the other calculations.
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie + User Effect Model",  
                                     RMSE = min(rmses)))
```

```{r table RMSE per method,echo=TRUE}
rmse_results %>% knitr::kable()
```


